---
title: "CIF"
author: "Yufang"
date: "2024-02-23"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Research question: Any causal relationship between education level and 
long-term (16 year) death.

# 1. The gobal DAG
```{r}
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)

g.postulated <- dagitty('dag {
  SEX  [pos="2.500,3.000"]
  AGE  [pos="-4.000,-1.000"]
  SYSBP  [pos="2.500,-1.000"]
  DEATH  [outcome,pos="-1.000,-2.000"]
  BPMEDS [pos="-1.000,0.000"]
  CURSMOKE [pos="-4.000,-3.000"]
  BMI [pos="-4.000,3.000"]
  HEARTRTE  [pos="-4.000,1.000"]
  DIABP  [pos="1.000,-3.000"]
  EDUC [exposure,pos="-1.000,3.000"]
  SEX   -> {DEATH SYSBP DIABP BPMEDS CURSMOKE EDUC HEARTRATE}
  AGE  -> { DEATH SYSBP DIABP BMI HEARTRATE EDUC}
  SYSBP  -> { BPMEDS}
  BMI -> DEATH
  EDUC -> BPMEDS
  HEARTRTE -> {SYSBP DIABP}
  }
')

## Plotting DAG
plot(g.postulated)
```

The minimal adjustment set is:
```{r, echo=TRUE}
adjustmentSets(g.postulated, type = "minimal", effect = "total")
```

# 3. Import the data and check the assumptions
```{r, echo=TRUE}
df = read.csv("framingham_assignment.csv")
df$SEX = factor(df$SEX)
df$SEX = factor(df$SEX, levels = c(1, 2), labels = c("Male", "Female"))
df$EDUC = factor(df$EDUC, levels = c(0, 1), labels = c("HighSchoolorBelow", "CollegeorMore"))
df$DEATH = factor(df$DEATH, levels = c(0, 1), labels = c("Alive", "Death"))

table(df$EDUC, df$DEATH)
mean(df$AGE[df$DEATH])
prop.table(table(df$EDUC, df$SEX), margin = 1)

library(vioplot)
vioplot::vioplot(df$AGE ~ df$EDUC, col = 2:3, xlab = "", ylab = "AGE")
```




# 4. Check the missing data issues
```{r, echo=TRUE}
ExpDf = data.frame(EDUC = df$EDUC, DEATH = df$DEATH, SEX = df$SEX, AGE = df$AGE)
colSums(apply(ExpDf, 2, is.na))
```
As shown in the table, Only data in variable 'EDUC' is missing. In our study,
variable 'EDUC' is the predictor instead of the response variable. Therefore, we
argue that the data set in this study is missing at random (MAR).

# 5. Statistical method
To obtain the causal conclusion from observational data, i.e., the case
of the current study, we picked up the G-computation (outcome regression + 
standardization) and the propensity score method (inverse probability weighing,
matching, stratification) to analysis data, as required by the assignment.

## outcome regression

In this analysis, we assume 1) conditional exchangeability, given age and gender as 
start, and 2) the linear model fits.
```{r, echo=TRUE}
model = glm(DEATH ~ SEX * EDUC*AGE, family = binomial(link = "logit"), data = ExpDf)
summary(model)

# Obtain the EY1hat and EY2hat
newdata = ExpDf
newdata$EDUC = "CollegeorMore"
Y1hat = predict(model, newdata = newdata, type = "response")


newdata$EDUC = "HighSchoolorBelow"
Y0hat = predict(model, newdata = newdata, type = "response")
mean(Y1hat) - mean(Y0hat)

library(stdReg)
fit.std = stdGlm(fit = model, data = ExpDf, X = "EDUC")
summary(fit.std, contrast = "difference", reference = "HighSchoolorBelow")
```
Report the result at the probability scale already.

### Assumption checking
```{r, echo=TRUE}
plot(model)
```
## the propensity score method (inverse probability weighing,
matching, stratification)
```{r, echo=TRUE}
ExpDf2 = ExpDf[complete.cases(ExpDf), ]
model2 = glm(EDUC ~ SEX + AGE, family = binomial(link = "logit"), data = ExpDf2)


ExpDf2$ps1 = fitted.values(model2)
range(ExpDf2$ps1)


ExpDf2$ipw1 <- (ExpDf2$EDUC == "CollegeorMore")/ExpDf2$ps1 + (ExpDf2$EDUC == "HighSchoolorBelow")/(1 - ExpDf2$ps1)
summary(ExpDf2$ipw1)

# c: checking the balance
vars1 = c("SEX", "AGE")
covariates = ExpDf2[, vars1]

library(cobalt)
bal.tab(covariates, treat =ExpDf2$EDUC, weights = ExpDf2$ipw1, method = "weighting", un = TRUE)

love.plot(
  covariates,
  treat = ExpDf2$EDUC,
  weights = ExpDf2$ipw1,
  method = "weighting",
  binary = "std",
  threshold = .1
)

#histogram
par(mfrow=c(2,1))
hist(ExpDf2$ps1[ExpDf2$EDUC == "CollegeorMore"], xlim=c(0,1),main="CollegeorMore", xlab="Propensity score")
hist(ExpDf2$ps1[ExpDf2$EDUC == "HighSchoolorBelow"], xlim=c(0,1),main="HighSchoolorBelow", xlab="Propensity score")


# f
ExpDf2$DEATH.num = as.numeric(ExpDf2$DEATH) - 1

# g
EY1 = weighted.mean(ExpDf2$DEATH.num[ExpDf2$EDUC == "CollegeorMore"], ExpDf2$ipw1[ExpDf2$EDUC == "CollegeorMore"])
EY0 = weighted.mean(ExpDf2$DEATH.num[ExpDf2$EDUC == "HighSchoolorBelow"], ExpDf2$ipw1[ExpDf2$EDUC == "HighSchoolorBelow"])
EY1 - EY0

# h
# ipw2.p99 = quantile(ExpDf2$ipw1, 0.99)
# ExpDf2$ipw1.trunc = ifelse(ExpDf2$ipw1> ipw2.p99, ipw2.p99, ExpDf2$ipw1)
library(survey)

d.w = svydesign(~ 1, weights = ExpDf2$ipw1, data = ExpDf2)
fit.ipw = svyglm(DEATH.num ~ EDUC, design = d.w)
summary(fit.ipw)

confint(fit.ipw)
```